---
title: "Adding new functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding new functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, echo=FALSE}
library(chefStats)
library(data.table)
library(kableExtra)
```

To add new functions to {chefStats}, you must follow these four general steps:


  1. Consider what type of function you need: `stat_by_strata_by_trt`, `stat_by_strata_across_trt` or `stat_across_strata_across_trt`. See article [Function types](https://hta-pharma.github.io/chefStats/articles/function_types.html.
  2. Decide what information the function needs in order to compute the desired results. See section [Interface with {chef}](#interface-with-chef)
  3. Write the function definition (use `use_chefStats()` to help create the template)
  4. Write the appropriate unit-tests for the function (using [testthat](https://testthat.r-lib.org/) framework)
<br>

### Walkthrough example

Here we show how we would add the function `n_subj_event()`, which counts the number of subjects experiencing the event by stratification level and treatment level, if it did not already exist in {chefStats}.

  1. Since the function produces a number by stratification level and by treatment level, this will be a `stat_by_strata_by_trt`
  2. The function will need to know. See section [Interface with chef](#interface-with-chef) for more details on how to pass this information from {chef} to {chefStats}:
      - Who was eligible to have the event
      - Who has the event
      - What the stratification "group" (e.g. SEX, AGE, etc) and level ("MALE", ">65", etc.) is
      - What the treatment level is (e.g. "TreatmentA")
  3. The function definition might look like this:
  
```{r, eval=FALSE}
  n_subj_event <-
  function(dat,
           event_index,
           cell_index,
           subjectid_var,
           ... # the `...` are required for all {chefStats} functions to "collect" any unused arguments passed from {chef}
           ) {
    # Please see the "Interface with {chef}" section for details on what
    # `event_index` and `cell_index`
    
    # `intersect()` provides us with a vector of rows in `dat` that match both
    # `event_index` and `cell_index` - aka records that were BOTH eligible to
    # have the event (`cell_index`) AND had the event (`event_index`)
    index <- intersect(event_index, cell_index)
    
    # Return all matching rows in `dat` where `INDEX_`
    # matches `index`.
    event_rows <- dat[INDEX_ %in% index]
    
    # `dat` contains event data, meaning subjects can appear more than once if
    # they have >1 event, so we need to remove these extra rows to get a proper
    # count
    event_rows_unique_by_subject <- unique(event_rows, by = subjectid_var)
    
    stat <- NROW(event_rows_unique_by_subject)

    # The return object has to be a data.table object with the following 3
    # columns. The `value` column always has to be a double (not an integer)
    return(
      data.table(
        description = "Number of subjects with events",
        label = "n_subj_events",
        value = as.double(stat)
      )
    )

  }


```
  
<style>
/* Remove border around tab content */
.tab-content {
    border: none !important;
}
</style>

<br>

### Interface with {chef} {.tabset .tabset-pills}

The statistical functions from {chefStats} will be called within the context of a {chef} pipeline. The category of the function determines what arguments are passed from {chef} to the function, however some arguments are passed to all categories.

The following table describes arguments that are passed to **all** {chefStats} function:

```{r, echo=FALSE}
dt <-
  data.table::data.table(
    `Argument Names` = c(
      "`dat`",
      "`event_index`",
      "`strata_var`",
      "`treatment_var`",
      "`subjectid_var`"
    ),
    Description = c(
      "A `data.table` containing the analysis data set produced by the `prepare_data` function. To allow flexability for creative use of {chefStats} functions, this dataset is **not** filtered to the exact records needed for each analysis when passed to {chefStats}. Instead this filtering is done inside the {chefStats} functions. This is done using the `INDEX_` column from `dat` that serves as a row ID, and is used for filtering with, for example, `cell_index` or `event_index`",
      "A `vector` of indicies indicating which rows (as specified in `INDEX_` column of `dat`) are considered to be events for the endpoint specification under evaluation",
      "A `character` indicating which stratatification is being used (e.g. SEX, AGE, etc)",
      "A `character` indicating the name of the column in `dat` containing the treatment information used for the endpoint",
      "A `character` specifying the name of the column in `dat` containing the subject ID. Defaults to \"USUBJID\""
    )
  )
dt |>
  kable(format = "html",
        table.attr = "class='table table-bordered'",
        caption = "Arguments always passed to {chefStats} functions") |>
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive")) |>
  row_spec(0, background = "lightblue")
```

<br><br>

Additionally, each function type receives the following arguments

#### stat_by_strata_by_trt

```{r, echo=FALSE}
dt <-
  data.table::data.table(
    `Argument Names` = c(
      "`cell_index`",
      "`strata_val`",
      "`treatment_val`"
    ),
    Description = c(
      "A `vector` of indicies specifying which rows in `dat` are considered to be part of the analysis for the given strata level and treatment level under evaluation. For example, if the current instance of the function was analysis \"Number of Events\" for SEX==\"M\" and TRT01A == \"Placebo\", then `cell_index` would be a vector of records in `dat$INDEX_` that match those parameters. You can thus obtain the analysis set by filtering `dat` via: `dat[cell_index %in% INDEX_]`",
      "A `character` specifying the stratification level under evaluation. For example if `strat_var ==\"SEX\"`, then `strat_val` could be either `\"M\"` or `\"F\"`",
      "A `character` specifying the treatment level (or treatment arm). *Not to be confused with the `treatment_refval` that specifies the reference treatment value."
    )
  )
dt |>
  kable(format = "html",
        table.attr = "class='table table-bordered'",
        caption = "Additional arguments passed to by_strata_by_trt functions") |>
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive")) |>
  row_spec(0, background = "lightblue")
```


#### stat_by_strata_across_trt


```{r, echo=FALSE}
dt <-
  data.table::data.table(
    `Argument Names` = c("`cell_index`",
                         "`strata_val`",
                         "`treatment_refval`"),
    Description = c(
      "A `vector` of indicies specifying which rows in `dat` are considered to be part of the analysis for the given strata level and treatment level under evaluation. For example, if the current instance of the function was analysis \"Number of Events\" for SEX==\"M\" and TRT01A == \"Placebo\", then `cell_index` would be a vector of records in `dat$INDEX_` that match those parameters. You can thus obtain the analysis set by filtering `dat` via: `dat[cell_index %in% INDEX_]`",
      "A `character` specifying the stratification level under evaluation. For example if `strat_var ==\"SEX\"`, then `strat_val` could be either `\"M\"` or `\"F\"`",
      "A `character` specifying the treatment reference level. *Not to be confused with the `treatment_val` that specifies the treatment value for `by_strata_by_trt` functions"
    )
  )
dt |>
  kable(format = "html",
        table.attr = "class='table table-bordered'",
        caption = "Additional arguments passed to by_strata_across_trt functions") |>
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive")) |>
  row_spec(0, background = "lightblue")
```
#### stat_by_strata_across_trt


```{r, echo=FALSE}
dt <-
  data.table::data.table(
    `Argument Names` = c("`strata_val`",
                         "`treatment_refval`"),
    Description = c(
      "A `character` specifying the stratification level under evaluation. For example if `strat_var ==\"SEX\"`, then `strat_val` could be either `\"M\"` or `\"F\"`",
      "A `character` specifying the treatment reference level. *Not to be confused with the `treatment_val` that specifies the treatment value for `by_strata_by_trt` functions"
    )
  )
dt |>
  kable(format = "html",
        table.attr = "class='table table-bordered'",
        caption = "Additional arguments passed to across_strata_across_trt functions") |>
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive")) |>
  row_spec(0, background = "lightblue")
```



<br>

### Using building-blocks

When possible, utilize building-block functions when making new statistical functions. For example, if the new functions requires a 2x2 table, use the `make_two_by_two_()` function instead of writing a new one.

This also allows you to easily write functions that collapse several {chefStats} functions into one function call. For example, on call to `count_set()` is the same as one call each to `n_sub()`, `n_event()`, `n_subj_event()` and `p_subj_event()`. The only rational for combining functions like this is to save compute time, due to the way {chef} pipelines are constructed. 

Building block function names are always suffixed with an underscore `_` to indicate they cannot be called from inside a {chef} pipeline. For example, `n_event_()` is a building block that is used to make `n_event()`, but `n_event_()` can also be use to build other functions, such as `count_set()`, because it does not format it's output for a {chef} pipeline. Conversely, `n_event()` does format the output, so it can be use in a {chef} pipeline, but not as a building block.
